\hl{UPDATE FROM WEDNESDAY}
\begin{itemize}
    \item Minimum papers to read for work leftover:
    \item CNNs 1 day
    \item HuBERT 1 day
    \item 2-3 neural scaling papers 1 day
    \item \textbf{To read but not quite as detailed.} Maybe 1 day for all of these.
    \item Whisper
    \item mel-cepstrum
    \item deepspeech and deepspeech 2
\end{itemize}


\begin{itemize}
    \item CNN - 2-3 pages
    \item CTC -  1-2 pages
    \item C/W/PER - 1-2 pages
    \item deepspeech - 2-3 pages
    \item spectrograms - 0.5-1 page
    \item LSTMs and RNNs - are they necessary?
    \item masking/specaugment - 0.5-1 pages (1 page with figure)
    \item layer freezing - 1 page
    \item how much of this is in deepspeech/deepspeech 2?
    \item auto-encoder - 0.5 pages
    \item activation functions ReLU/softmax etc. 
    \item normalisation 
    \item backpropogation - 0.5 pages at most
    \item learning rates? - 0.5 but could integrate with backpropogation
    \item how much detail do you actually need to go to?
    \item Articulations and linguistics
    \item ask sam about some of this
    \item \hl{Other sections}
    \item transformer
    \item conformer
    \item HuBERT 
    \item neural scaling
\end{itemize}

How much of this do I actually need?
\begin{itemize}
    \item transformer - 2 pages
    \item model scaling - 2 pages
    \item \textbf{Short sections 0.5-1 pages}
    \item HuBERT, whisper conformer - page for all
    \item MFA?
    \item CTC - 0.5 pages max
    \item WER/CER/PER - 0.5 pages max
    \item \textbf{Section to combine 1 page max for all of these things}
    \item Deepspeech 1/2
    \item backpropogation
    \item learning rate
    \item activation functions
    \item spectrograms
\end{itemize}

to do:
1-2 pages on transformers whisper BERT HuBERT and WavLM
1 pages on neural scaling
2 pages on historical ASR.

Introduction?
Clean everything up.

Sam's icassp
\begin{itemize}
    \item lots of evaluation papers
    \item soundhound 
    \item a lot of talk about agentic AI and how models are hallucinating and ding things like giving refunds for stuff it shouldn't 
    \item recruiters:
    \begin{itemize}
        \item researcher role is becoming less prominent
        \item Look for more development engineers
        \item Sam's leaning on pre-PhD private sector experience
        \item Competing against people who have industry experience
        \item companies going into hiring freezes
    \end{itemize}
    \item LinkedIn is the most common place to get interviews
    \item leetcode is still a big thing
    \item 
\end{itemize}